---
title: "data_manipulation"
output: github_document
date: "2025-09-18"
---

```{r}
library(tidyverse)
```

Let's import the datasets.

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", ""))
pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", na = c("NA", ".", ""), skip=3)

litters_df =
  janitor::clean_names(litters_df)
pups_df =
  janitor::clean_names(pups_df)
```

## `select`

`select` selects variables from my dataframe.

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range

```{r}
select(litters_df, group:gd_of_birth)
```

select by removal

```{r}
select(litters_df, -group)
select(litters_df, -group:gd_of_birth)
```

select by prefix

```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting

```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())

relocate(litters_df, litter_number) #does the same thing
```

rename by selecting

```{r}
select(litters_df, GROUP = group, everything())
rename(litters_df, GROUP = group) #does same thing
```

what if i only select one variables?

```{r}
select(litters_df, group) #produces df

pull(litters_df, group) #produces a vector (almost never use)
```

select litter number, sex, PD ears from pups dataset

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## `filter`

use `filter()` to filter out rows

```{r}
filter(litters_df, gd_of_birth == 20, pups_born_alive == 6)
```

look for more than five pups alive

```{r}
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive >= 5)
filter(litters_df, pups_born_alive <= 5)
```

look for non-matches

```{r}
filter(litters_df, pups_born_alive != 6)
```

match to a set

```{r}
filter(litters_df, group %in% c("Con7", "Con8"))
```

filter out missing rows with `drop_na()`

```{r}
drop_na(litters_df)
drop_na(litters_df, gd0_weight)
```

filter to include pups with sex 1, and only pups with PD walk less than 11 and sex 2

```{r}
filter(pups_df, sex == 1)
filter(pups_df, pd_walk < 11, sex == 2)
```

